Congestion-Aware Join Optimization in Key-Value Stores

### 2014-08-07 paper/paper.tex
Adds initial version of the evaluation

As we discuss later, some existing query processing engines take into account network topology and link utilization when scheduling distributed queries.
However, there is relatively little work expoloiting dynamic reconfiguration of the network to support query processing.
A centralized network controller which is aware of the dynamics of query processing is able to make smart routing decisions to optimize performance.
For example, flows for queries which are latency sensitive can be given priority over highly utilized links, while other flows can be routed to longer paths with lower utilization.
This can be done dynamically based on the state of the network.
Furthermore, the routing can be done using hardware switches as opposed to intelligent software routers, which minimizes latency and improves scalability.

@@

The execution model for these queries is very straightforward.
We consider the execution of the simple query given above.
We start by passing the query to each node which has entities of the class on the left side of the join (\texttt{Item}).
Those nodes will scan their local list of entities and apply the necessary filters (\texttt{price > 20}).
Each node then forwards the necessary foreign keys to nodes holding entites of the class on the right side of the join (\texttt{User}).
These nodes are then able to look up entities on the other side of the join and results can be passed back to the client.

@@

Simple approaches to bandwith allocation such as max-min fairness pose significant problems in our settings.
If we were to use fair share allocation, then the node with the largest share of the query results, bottlenecked by available bandwidth, would become a straggler.
Nodes with a small share of query results could finish much more quickly only to wait for these stragglers.
To minimize the overall completion time of queries, we can allocate bandwidth to each node as the fraction of query results which will be returned by the node.
This will speed up stragglers at the expense of slowing down nodes with a smaller allocation.
However, the net result is a reduction in the average execution cost.

@@

CloudTPS~\cite{Wei2012} provides a join query processing layer on top of NoSQL stores which enforces ACID semantics for transactions.
Our system does not currently consider transactional consistency, so this could be an interesting area to explore.
R\"{o}diger et al.~\cite{Rodiger2014} examine the affects of data locality with respect to distributed query processing.
Their Neo-Join algorithm is designed for efficient network usage by repartitioning data and scheduling network transfers to avoid cross-traffic.
These techniques exert no control over the network, which may provide further opportunities for optimization.

For example, Xiong et al.~\cite{Xiong2014} examine the usefulness of software-defined networking in supporting distributed analytical queries.
Their workload consits of read-only SQL queries where query processing is bandwidth-intensive.
They construct a global query optimizer which decides on join order and how query results should be passed between sites.
This work takes a similar approach to ours, but also identifies many opportunities for further work.

@@

@@

There are many opportunities to combine this work with appropriate key placement strategies for further performance optimizations.
For heterogenous networks, a key placement strategy which is network and workload aware may provide significant performance advantages.
Our current query execution model is very simplistic and could be expanded to support richer queries and distributed index data structures.
We would also like to investigate other query types such as aggregation which we suspect would also benefit from our approach.
This presents new challenges in network scheduling, especially in the face of concurrent queries of multiple different types.

>>>>

As we discuss later, some existing query processing engines take into account network topology and link utilization when scheduling distributed queries.
However, there is relatively little work exploiting dynamic reconfiguration of the network to support query processing.
A centralized network controller which is aware of the dynamics of query processing is able to make smart routing decisions to optimize performance.
For example, flows for queries which are latency sensitive can be given priority over highly utilized links, while other flows can be routed to longer paths with lower utilization.
This can be done dynamically based on the state of the network.
Furthermore, the routing can be done using hardware switches as opposed to intelligent software routers, which minimizes latency and improves scalability.

@@

The execution model for these queries is very straightforward.
We consider the execution of the simple query given above.
We start by passing the query to each node which has entities of the class on the left side of the join (\texttt{Item}).
Those nodes will scan their local list of entities and apply the necessary filters (\texttt{price > 20}).
Each node then forwards the necessary foreign keys to nodes holding entities of the class on the right side of the join (\texttt{User}).
These nodes are then able to look up entities on the other side of the join and results can be passed back to the client.

@@

Simple approaches to bandwidth allocation such as max-min fairness pose significant problems in our settings.
If we were to use fair share allocation, then the node with the largest share of the query results, bottlenecked by available bandwidth, would become a straggler.
Nodes with a small share of query results could finish much more quickly only to wait for these stragglers.
To minimize the overall completion time of queries, we can allocate bandwidth to each node as the fraction of query results which will be returned by the node.
This will speed up stragglers at the expense of slowing down nodes with a smaller allocation.
However, the net result is a reduction in the average execution cost.

@@

CloudTPS~\cite{Wei2012} provides a join query processing layer on top of NoSQL stores which enforces ACID semantics for transactions.
Our system does not currently consider transactional consistency, so this could be an interesting area to explore.
R\"{o}diger et al.~\cite{Rodiger2014} examine the affects of data locality with respect to distributed query processing.
Their Neo-Join algorithm is designed for efficient network usage by re-partitioning data and scheduling network transfers to avoid cross-traffic.
These techniques exert no control over the network, which may provide further opportunities for optimization.

For example, Xiong et al.~\cite{Xiong2014} examine the usefulness of software-defined networking in supporting distributed analytical queries.
Their workload consists of read-only SQL queries where query processing is bandwidth-intensive.
They construct a global query optimizer which decides on join order and how query results should be passed between sites.
This work takes a similar approach to ours, but also identifies many opportunities for further work.

@@

Each experiment is performed by 5 independent runs and the results are used to
compute the data in the following sections. The 95th percentile confidence
interval is shown whenever appropriate.
\subsection{Micro Benchmark}
\begin{figure}
    \centering
    \includegraphics[scale=2.5]{figures/Join_micro_benchmark.png}
    \caption{Micro benchmark setup}\label{fig:micro_benchmark}
\end{figure}
First, we want to demonstrate the proof-of-concept of our design by evaluating
our system with the simple topology shown in figure
\ref{fig:micro_benchmark}. We engineered the client query such that each server
holds a different portion of the relevant data for the join operation,
i.e. server 1 holds a quarter of the relevant \texttt{item} data and server 2,
3, 4 holds 1/16, 1/32, 1/64 of the relevant \texttt{item} data
respectively. The total relevant data sizes of \texttt{item} and \texttt{user}
are about 9 MB and 2 MB respectively. In order to satisfy our assumption where
the network is the bottleneck, we set the capacity of each link in the topology
to 1Mbps.  The client is connected directly to the core switch with the same
link capacity.

In figure \ref{fig:micro_benchmark_total_time}, the red bars represent our
system's baseline performance where the network traffic scheduling is turned
off, i.e. without SDN. The average completion time without scheduling for each
client query is over 180 seconds with a high variation. This kind of
unpredictable performance is undesired in any datacenters. To provide
predictable performance with our scheduling mechanism, we must first understand
the characteristics of the costs for each sub query of the join operation. In
our case it is easy to predict the costs because we assigned different portions
of relevant data to each server, however, it is also easy in real deployment
since the statistics of the server is usually available. Recall that by our
design, server 1 is responsible to transmit the largest flows from and to other
servers and the client.  Therefore, our scheduling must take this information
into consideration and ensure that server 1 have the largest minimum bandwidth
guarantee for this particular operation. Though we want server 1 to transfer
its data as fast as possible, we do not want to starve other
servers. Therefore, we must also have reasonable minimum bandwidth guarantees
for other servers. Through experimentation, we find the most suitable minimum
bandwidth guarantees on the 1 Mbps link for servers 2, 3, and 4 are 0.3 Mbps,
0.2 Mbps and 0.1 Mbps respectively. We fixed these parameters and varied the
minimum bandwidth guarantee for the server 1 to study the network
behaviours. As shown by the yellow bars in figure
\ref{fig:micro_benchmark_total_time}, there is noticeable performance gain when
we increase the minimum bandwidth guarantee of server 1 from 0.5 Mbps to 0.8
Mbps. It shows that it is important to favour the elephant flows during a join
operation to minimize the overall completion time. We also want to highlight
that we see diminished return as we increase the guarantee to 0.9 Mbps or even
1.0 Mbps because this may significantly hurt other flows' completion time which
result in slower execution overall. In the best case, with 0.8 minimum
bandwidth guarantee for server 1, the average completion time of the client
query is 154 seconds which gives us almost 15\% performance gain over the
baseline implementation.

\begin{figure}
    \centering
    \includegraphics[scale=0.47]{figures/simple_topo_overall.png}
    \caption{Query completion time}\label{fig:micro_benchmark_total_time}
\end{figure}

Next, we want to take a closer look at the performance results to understand
why network scheduling helps. Figure \ref{fig:sub_query_completion_8} shows the
sub query completion time for each server. The per server completion is
calculated by the client after it fetched all the required data from each
server. The white bars on the left is generated when the traffic scheduling is
turned off. We see that server 1 takes the most time to complete and server 4
takes the least time. This is expected due to the relevant data
distribution. We can also see that if we turned on our network traffic
scheduling mechanisms, the completion time for server 3 and server 4 slowed
down quite a bit. This is because most of the bandwidth capacity will be
allocated to help server 1 to finish its task as soon as possible to minimize
final query completion time. The end result is of traffic scheduling is to
average out the sub query completion time which in turn reduces the overall
query execution time. We see that our experimental results match with our
expectations.

\begin{figure}
    \centering
    \includegraphics[scale=0.47]{figures/simple_topo_closer_8.png}
    \caption{Sub query completion time}\label{fig:sub_query_completion_8}
\end{figure}

@@

There are many opportunities to combine this work with appropriate key placement strategies for further performance optimizations.
For heterogeneous networks, a key placement strategy which is network and workload aware may provide significant performance advantages.
Our current query execution model is very simplistic and could be expanded to support richer queries and distributed index data structures.
We would also like to investigate other query types such as aggregation which we suspect would also benefit from our approach.
This presents new challenges in network scheduling, especially in the face of concurrent queries of multiple different types.
